Learning to Dispatch: A Reinforcement Learning Framework for Train Dispatch Networks

### 2025-04-23 paper/main.tex
Locked in to background and introduction

In practice, the TDP is currently managed either by human dispatchers or by static rule-based systems. 
Human expertise plays a crucial role in adapting to disruptions or prioritizing certain services, but manual dispatching lacks scalability and consistency. 
Rule-based systems, while automated, are inflexible and often suboptimal when network conditions change. 
These limitations open the door for machine learning-based methods, particularly those capable of learning adaptive and scalable policies from data and interaction.

@@

Building on this foundation, Silver et al.~\cite{drl:silver2017mastering} showcased the power of DRL in mastering the game of Go, a task long considered a grand challenge in artificial intelligence. 
Their AlphaGo system employed deep neural networks to evaluate board positions and guide search strategies, achieving superhuman performance. 
Subsequently, the AlphaZero algorithm~\cite{drl:silver2017alphazero} extended this approach to generalize across multiple games, such as chess and shogi, using a self-play mechanism and a unified reinforcement learning framework.

These advancements have established DRL as a versatile and powerful tool for solving sequential decision-making problems in diverse domains, including robotics, healthcare, and transportation systems.
\subsubsection{Markov Decision Processes}
Markov Decision Processes (MDPs) provide a mathematical framework for modeling decision-making in environments where outcomes are partly random and partly under the control of a decision-maker. An MDP is defined by the tuple $(S, A, P, R, \gamma)$:

@@

@@

\begin{itemize}
    \item \textbf{Q-Network:} A neural network that takes the current state $s_t$ as input and outputs Q-values for all possible actions. The network is parameterized by weights $\theta$, which are updated during training.
    \item \textbf{Target Network:} A separate network with weights $\theta^-$, used to stabilize training by providing target Q-values. The target network is periodically updated to match the Q-network.
    \item \textbf{Experience Replay:} A buffer that stores past experiences $(s_t, a_t, r_t, s_{t+1})$. During training, mini-batches of experiences are sampled randomly from the buffer to break correlations and improve learning stability.
    \item \textbf{Bellman Equation:} The Q-network is trained to minimize the temporal difference (TD) error, defined as:
    \[
    \delta = \left[ r_t + \gamma \max_{a'} Q(s_{t+1}, a'; \theta^-) - Q(s_t, a_t; \theta) \right],
    \]
    where $\gamma$ is the discount factor.
    \item \textbf{Exploration-Exploitation Tradeoff:} The agent balances exploration (choosing random actions) and exploitation (choosing actions with the highest Q-value) using an $\epsilon$-greedy policy.
\end{itemize}

The training process involves iteratively updating the Q-network's weights $\theta$ using gradient descent to minimize the TD error. The learned Q-values guide the agent in selecting actions that maximize long-term rewards.

@@

\subsubsection{Overview of GNNs}
GNNs extend traditional neural networks by incorporating message-passing mechanisms, where nodes aggregate information from their neighbors iteratively. This process enables the network to capture both local and global graph structures. The key components of GNNs include:

\begin{itemize}
    \item \textbf{Message Passing:} Nodes exchange information with their neighbors through a series of message-passing steps. At each step, a node updates its representation by aggregating messages from its neighbors.
    \item \textbf{Aggregation Functions:} Common aggregation functions include summation, mean, and max-pooling, which ensure permutation invariance in the graph structure.
    \item \textbf{Update Functions:} After aggregation, node representations are updated using learnable functions, often implemented as neural networks.
    \item \textbf{Pooling Layers:} For graph-level tasks, pooling layers summarize the entire graph into a fixed-size representation.
\end{itemize}

@@

In multi-objective optimization problems, such as influence maximization in social networks, GNNs combined with meta-reinforcement learning have demonstrated effectiveness. For instance, the GraMeR framework~\cite{gnndrl:munikoti2022gramergraphmetareinforcement} leverages GNNs to learn transferable policies across different graph instances, optimizing multiple objectives simultaneously.

Devailly et al.~\cite{gnndrl:Devailly_2022} introduce IG-RL, an inductive graph reinforcement learning approach for large-scale traffic signal control. By using GNNs, their method efficiently handles massive traffic networks, enabling scalable and adaptive control strategies.

In summary, GNNs provide a powerful framework for learning on graph-structured data, with significant implications for reinforcement learning in domains like transportation, social networks, and beyond.

>>>>

In practice, the TDP is currently managed either by human dispatchers or by static rule-based systems. 
Human expertise plays a crucial role in adapting to disruptions or prioritizing certain services, but manual dispatching lacks scalability and consistency. 
Rule-based systems, while automated, are quite inflexible and often suboptimal when conditions change. 
These limitations open the door for machine learning-based methods, particularly those capable of learning scalable policies from data and interaction.

@@

These advancements have established DRL as a versatile and powerful tool for solving sequential decision-making problems in diverse domains, including robotics, healthcare, and transportation systems.
\subsubsection{Markov Decision Processes}
Markov Decision Processes (MDPs) form the backbone of reinforcement learning. 
MDPs offer a structured framework for modeling decision-making in environments where outcomes are influenced by both randomness and the agent's actions. 
An MDP is formally defined by the tuple $(S, A, P, R, \gamma)$, where:

@@

In essence, an MDP is a process where the current state contains all the information necessary to determine the optimal action and its associated reward. 
In reinforcement learning, this process is typically modeled as an interaction between an agent and an environment, governed by underlying dynamics. 
At each timestep, the environment provides the current state $s_t$ to the agent, which selects an action $a_t$. 
This action is then applied to the environment, which updates its state to $s_{t+1}$ and computes the corresponding reward $r_{t+1}$. 
This iterative process is illustrated in Figure~\ref{fig:rl_framework}.

@@

\begin{itemize}
    \item \textbf{Q-Network:} A neural network that takes the current state $s_t$ as input and outputs Q-values for all possible actions. 
    \item \textbf{Target Network:} A separate network with weights $\theta^-$, used to stabilize training by providing target Q-values. The target network is periodically updated to match the Q-network.
    \item \textbf{Experience Replay:} A buffer that stores past experiences $(s_t, a_t, r_t, s_{t+1})$. During training, mini-batches of experiences are sampled randomly from the buffer to break correlations and improve learning stability.
    \item \textbf{Bellman Equation:} The Q-network is trained to minimize the temporal difference (TD) error, defined as:
    \[
    \delta = \left[ r_t + \gamma \max_{a'} Q(s_{t+1}, a'; \theta^-) - Q(s_t, a_t; \theta) \right],
    \]
    where $\gamma$ is the discount factor.
    \item \textbf{Exploration vs. Exploitation:} The agent balances exploration (choosing new actions to explore the environment) and exploitation (choosing actions to exploit what it already knows) using an $\epsilon$-greedy policy.
\end{itemize}

The training process involves iteratively updating the Q-network's weights $\theta$ using gradient descent to minimize the TD error. 
The learned Q-values guide the agent in selecting actions that maximize long-term rewards.
For simpler tasks, a basic Q-network may suffice (such as a linear regression or heuristic model). 
However, for complex problems like the TDP, more advanced neural network architectures are likely better suited to handle the challenges effectively.

@@

GNNs extend traditional neural networks by incorporating message-passing mechanisms, where nodes aggregate information from their neighbors iteratively. 
This process enables the network to capture both local and global graph structures. 
The key components of GNNs include:

\begin{itemize}
    \item \textbf{Message Passing:} Nodes exchange information with neighbors through a series of message-passing steps. At each step, a node updates its representation by aggregating messages from its neighbors.
    \item \textbf{Aggregation Functions:} Common aggregation functions include summation, mean, and max-pooling.
    \item \textbf{Update Functions:} After aggregation, node representations are updated using learnable functions, often implemented as neural networks.
    \item \textbf{Pooling Layers:} For graph-level tasks, pooling layers summarize the entire graph into a fixed-size representation.
\end{itemize}

Graph Neural Networks (GNNs) can be thought of as the graph-based counterpart to Convolutional Neural Networks (CNNs) for images. 
GNNs effectively aggregate information from neighboring nodes and edges, enabling each node to build a representation that incorporates both its own features and the context provided by its local graph structure.
They then apply a pooling operation to extract and summarize key patterns and trends from the data.
This comparison is illustrated in Figure~\ref{fig:cnn_vs_gnn}

\begin{figure}[htbp]
    \centering
    % CNN
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \begin{tikzpicture}[
            image/.style={draw, rounded corners, fill=orange!30, minimum width=2cm, minimum height=2cm},
            filter/.style={draw, thick, fill=green!20, minimum width=1cm, minimum height=1cm, rounded corners},
            feature/.style={draw, thick, fill=green!20, minimum width=1cm, minimum height=0.6cm, rounded corners},
            ann/.style={draw, thick, fill=purple!30, circle, minimum size=0.8cm},
            arrow/.style={->, thick, >=Stealth},
            every node/.style={font=\scriptsize}
        ]

        % Nodes
        \node[image] (input) at (0, 0.25) {\includegraphics[width=1.5cm]{example-image}};
        \node[filter] (filter) at (0, -1.75) {Filter};
        \node[feature] (feature) at (0, -3) {Pooling};
        \node[ann] (ann) at (0, -4.5) {NN};

        \end{tikzpicture}
        \caption{Convolutional Neural Network}
    \end{subfigure}
    \hfill
    % GNN
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \begin{tikzpicture}[
            node/.style={circle, draw, fill=orange!30, minimum size=0.8cm},
            message/.style={draw, thick, fill=blue!20, minimum width=1cm, minimum height=0.6cm, rounded corners},
            aggregate/.style={draw, thick, fill=green!20, minimum width=1cm, minimum height=0.6cm, rounded corners},
            arrow/.style={->, thick, >=Stealth},
            doublearrow/.style={<->, dotted, thick, >=Stealth},
            every node/.style={font=\scriptsize}
        ]

        % Nodes
        \node[node] (A) at (0, 0) {A};
        \node[node] (B) at (1.5, 1) {B};
        \node[node] (C) at (1.5, -1) {C};
        \node[node] (D) at (3, 0) {D};

        % Message Passing
        \node[message] (msg) at (1.5, -2) {Message Passing};
        \draw[arrow] (C) -- (msg);
        \draw[arrow] (A) -- (msg);
        \draw[arrow] (D) -- (msg);

        % Aggregation
        \node[aggregate] (agg) at (1.5, -3) {Aggregation};
        \draw[arrow] (msg) -- (agg);

        % Pooling
        \node[aggregate] (pooling) at (1.5, -4) {Pooling};
        \draw[arrow] (agg) -- (pooling);

        % Output
        \node[node, fill=purple!30, minimum size=0.8cm, circle] (output) at (1.5, -5) {NN};
        \draw[arrow] (pooling) -- (output);

        \end{tikzpicture}
        \caption{Graph Neural Network}
    \end{subfigure}
    \caption{Comparison of CNN and GNN architectures.}
    \label{fig:cnn_vs_gnn}
\end{figure}

@@

Graph Neural Networks (GNNs) have been successfully integrated with reinforcement learning (RL) to address complex graph-structured problems across various domains. 
For example, Almasan et al.~\cite{gnndrl:ALMASAN2022184} applied GNNs to network routing, demonstrating their ability to optimize routing decisions in dynamic environments.
Tackling the problem of influence maximization in social networks, GraMeR~\cite{gnndrl:munikoti2022gramergraphmetareinforcement} leverage GNNs combined with meta-reinforcement learning to learn transferable policies across diverse graph instances.

Devailly et al.~\cite{gnndrl:Devailly_2022} introduced IG-RL, an inductive graph reinforcement learning approach for large-scale traffic signal control. 
By utilizing GNNs, IG-RL efficiently manages massive traffic networks, enabling scalable and adaptive control strategies. 
Similarly, Agasucci et al.~\cite{gnndrl:agasucci} proposed a DRL-based formulation for the train dispatching problem, although their approach is constrained by the specificity of the data source used.

In summary, GNNs offer a robust framework for learning on graph-structured data, significantly augmenting the set of possible reinforcement learning applications.
The ability of GNNs to learn from dynamic graph-structured data is a core concept behind this formulation.
